---
title: "Analyzing Maryland Campaign Finance Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analyzing Maryland Campaign Finance Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

`freestatemoney` downloads and parses campaign finance data from the Maryland
State Board of Elections Campaign Reporting Information System (CRIS). This
vignette walks through the full workflow: getting data, loading it, and
answering common reporting questions.

```{r setup}
library(freestatemoney)
library(dplyr)
```

## Getting data

`md_download()` fetches bulk CSVs directly from the SBE's API — the same data
the download page at
<https://campaignfinance.maryland.gov/public/cf/downloads> serves. Full-cycle
transaction files can run to hundreds of megabytes, so pass a `year` to limit
downloads to a single filing year:

```{r download, eval = FALSE}
committees <- md_committees(md_download("committees"))
contributions <- md_contributions(md_download("contributions", year = 2025))
expenditures <- md_expenditures(md_download("expenditures", year = 2025))

# Or all three at once:
md <- md_load_all(year = 2025)
```

So this vignette can run anywhere, it uses small samples of real CRIS data
that ship with the package:

```{r load-samples}
sample_file <- function(name) {
  system.file("extdata", name, package = "freestatemoney")
}

committees <- md_committees(sample_file("committees_sample.csv"))
contributions <- md_contributions(sample_file("contributions_sample.csv"))
expenditures <- md_expenditures(sample_file("expenditures_sample.csv"))
```

## What the loaders handle for you

Raw CRIS files have several quirks that the loaders take care of:

- A metadata line like `Committee Download as of 06/14/2026 01:00 AM`
  precedes the header. It is skipped, and its timestamp is kept:

```{r download-date}
attr(committees, "download_date")
```

- Amounts arrive dollar-formatted (`$3,000.00`) and are parsed as numeric;
  dates (`MM/DD/YYYY`) become `Date` objects:

```{r types}
contributions %>%
  select(transaction_date, transaction_amount) %>%
  head(3)
```

- Zip codes arrive wrapped for Excel (`="21228"`) and are unwrapped; files
  are read with their actual encoding (typically Windows-1252, not UTF-8).

The loaders also refuse to load the wrong dataset — passing a contributions
file to `md_committees()` produces an error naming the function you wanted.

## Exploring committees

```{r committees}
committees %>%
  count(committee_type, sort = TRUE)

committees %>%
  filter(committee_type == "Candidate Committee") %>%
  select(committee_name, candidate_last_name, office_sought, party_affiliation) %>%
  head(5)
```

## Who gives, and how much?

`md_top_contributors()` aggregates by contributor — organizations by company
name, individuals as "Last, First" — and ranks by total:

```{r top-contributors}
top <- md_top_contributors(contributions, n = 10)
top
```

```{r top-contributors-plot, fig.alt = "Horizontal bar chart of total contributions by contributor from the package sample data"}
library(ggplot2)

top %>%
  mutate(contributor = reorder(contributor, total)) %>%
  ggplot(aes(x = total, y = contributor)) +
  geom_col(fill = "#4666DE", width = 0.7) +
  scale_x_continuous(labels = scales::dollar) +
  labs(
    title = "Total contributions by contributor",
    subtitle = "Package sample data",
    x = NULL, y = NULL
  ) +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank())
```

With a full year of data, the same call surfaces the largest donors in the
state.

## Following the spending

```{r spending}
expenditures %>%
  group_by(category) %>%
  summarize(
    total = sum(transaction_amount, na.rm = TRUE),
    n = n()
  ) %>%
  arrange(desc(total))
```

Independent expenditures and electioneering communications — spending *about*
candidates rather than *by* them — get their own summary. CRIS labels these
with combined transaction types, so `md_independent_expenditures()` matches
by pattern:

```{r ie}
md_independent_expenditures(expenditures)
```

## Committee-level totals

`md_committee_summary()` joins all three datasets on `filing_entity_id` and
totals each committee's fundraising and spending. Outstanding obligations
(unpaid debts) are excluded from spending, and committees with no
transactions get zeros:

```{r summary}
md_committee_summary(committees, contributions, expenditures) %>%
  head(5)
```

Note that the sample files cover different committees, so totals here are
mostly zero; with real full-year downloads every active committee is
represented. `net_raised` reflects only the transactions in the data you
loaded — it is not the committee's official cash balance.

## Linking datasets yourself

`filing_entity_id` links all three datasets, so any join works:

```{r join, eval = FALSE}
contributions %>%
  left_join(
    committees %>% select(filing_entity_id, office_sought, party_affiliation),
    by = "filing_entity_id"
  ) %>%
  group_by(party_affiliation) %>%
  summarize(total_raised = sum(transaction_amount, na.rm = TRUE))
```

## Data notes

- **Lump-sum contributions** have blank contributor details; they are dropped
  by `md_top_contributors()` but present in the raw data.
- **Aggregate As Of Download Date** gives each contributor's running total to
  that committee as of the file's extraction date.
- Contributor names are as reported: the same person may appear under
  multiple spellings.
- The SBE regenerates bulk files roughly daily; check
  `attr(df, "download_date")` for your file's extraction time.
