---
title: "Working with Large Files via DuckDB"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with Large Files via DuckDB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
has_duckdb <- requireNamespace("duckdb", quietly = TRUE) &&
  requireNamespace("DBI", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = has_duckdb
)
```

Full-cycle CRIS transaction files are large — a single year of contributions
runs to hundreds of megabytes — and multi-year analyses multiply that. The
`md_db()` family ingests bulk data into a persistent
[DuckDB](https://duckdb.org/) database instead of an in-memory tibble: you
download and parse once, query lazily with dplyr, and only your results ever
come into R.

These functions need the suggested packages duckdb and DBI
(`install.packages(c("duckdb", "DBI"))`).

```{r setup}
library(freestatemoney)
library(dplyr)
```

## Building a database

`md_db()` opens (or creates) a database file; `md_db_load()` downloads a
dataset and ingests it. Different years accumulate; re-ingesting a year
replaces it, so refreshes are idempotent:

```{r build, eval = FALSE}
con <- md_db("maryland.duckdb")

md_db_load(con, "committees")
md_db_load(con, "contributions", year = 2024)
md_db_load(con, "contributions", year = 2025)
md_db_load(con, "expenditures", year = 2025)
```

Because the database file persists, later sessions just reopen it — no
re-downloading:

```{r reopen, eval = FALSE}
con <- md_db("maryland.duckdb")
md_db_tables(con)
```

For this vignette we build a throwaway database from the sample data shipped
with the package:

```{r sample-db}
sample_file <- function(name) {
  system.file("extdata", name, package = "freestatemoney")
}

con <- md_db(tempfile(fileext = ".duckdb"))
md_db_load(con, "committees", file = sample_file("committees_sample.csv"))
md_db_load(con, "contributions", file = sample_file("contributions_sample.csv"))
md_db_tables(con)
```

## Querying with dplyr

`dplyr::tbl()` gives a lazy table: filters, joins, and aggregations compile
to SQL and run inside DuckDB. Data only enters R when you `collect()`:

```{r query}
contributions <- tbl(con, "contributions")

contributions |>
  group_by(contributor_type) |>
  summarize(
    total = sum(transaction_amount, na.rm = TRUE),
    n = n()
  ) |>
  arrange(desc(total)) |>
  collect()
```

Joins across datasets work the same way, entirely inside the database (the
two sample files happen to cover different committees, so this join is
empty here; with real full-year downloads every committee matches):

```{r join}
tbl(con, "contributions") |>
  inner_join(
    tbl(con, "committees") |> select(filing_entity_id, office_sought),
    by = "filing_entity_id"
  ) |>
  count(office_sought) |>
  collect()
```

The `filing_year` column added at ingest lets multi-year analyses group by
year:

```{r by-year, eval = FALSE}
tbl(con, "contributions") |>
  group_by(filing_year) |>
  summarize(total_raised = sum(transaction_amount, na.rm = TRUE)) |>
  collect()
```

## Parsing parity with the loaders

Ingestion applies the same treatment as `md_contributions()` and friends:
the metadata line is skipped, names are cleaned to snake_case, dates and
dollar amounts are typed, Excel armor is stripped, and CRIS export defects
(bare line breaks inside fields, Windows-1252 bytes, truncated or over-long
records) are repaired identically. A query against the database returns the
same values a tibble would:

```{r parity}
tbl(con, "contributions") |>
  select(transaction_date, transaction_amount, contributor_zip_code) |>
  head(3) |>
  collect()
```

## Cleaning up

```{r cleanup}
DBI::dbDisconnect(con)
```

The `.duckdb` file remains on disk for next time (delete it like any file if
you're done with it).
